---
title: "Project 2"
author: "Yvette Callender"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages used in this project

```{r, message=FALSE}
library(tidyverse)
#library(httr)
library(jsonlite)
```

# Introduction

The API chosen for investigation was data from the Center for Disease Control relating to [Vaccination Coverage among Adolescents](https://data.cdc.gov/Teen-Vaccinations/Vaccination-Coverage-among-Adolescents-13-17-Years/ee48-w5t6).  The dataset contains columns for vaccine, dose, geography type (national, regional, state, and local area), geography name, survey year, dimension type, dimension,estimate 95%CI, and sample size.  Dimension types include age, race/ethnicity, insurance coverage, poverty levels, proximity to an urban area, and overall.  This suggests areas of inquiry regarding things like:  

- geographical variation in vaccination coverage at regional level or state level  
- geographical variation in vaccination coverage differences due to  race/ethnicity or insurance coverage at the 
regional level and state level  
- variation in vaccination coverage based on the vaccine in question

The vaccines covered by the database include:

- HPV (human papillomavirus)  
- Menigococcal Conjugate  
- T or Td  (tetanus or tetanus/diptheria)  
- Tdap (tetanus, diptheria, and pertussis) (Replaced Td in 2005) 
- Varicella (chicken pox)  
- Hep A (hepatitis A)  
- MMR (measles, mumps, rubella)  
- HepB (hepatitis B)  

Parental attitudes may vary a great deal between these vaccines due to several factors such as perception of the threat being protected against (chicken pox not as serious as diptheria), whether the threat is associated with sexual behavior or drug use (HPV or HepB), and whether the vaccine has received particular attention from the anti-vaccination activists (MMR).  The HPV vaccine may also be perceived as more valuable to children who are assigned female at birth than those assigned male at birth. 

For HPV,the database uses the dose variable as way to track differential treatment of "male" and "female" children.

In this project, functions were developed to help users query the dataset about the following:

- HPV Vaccination Coverage: Regional level differences between genders with the option of changing dimension (race/ethnicity, poverty, insurance coverge, urbancity)  
- HPV Vaccination Coverage: Regional level differences in overall coverage estimate with the option of changing dimension (race/ethnicity, poverty, insurance coverage, urbancity)  
- Vaccination for Vaccines other than HPV: Regional level differences in coverage estimates with the option of changing vaccine and dimension  
- Vaccination for Vaccines other than HPV: State level differences in coverage estimates with the option of changing vaccine and dimension.  

Geographical information in the database is presented at a state level and at a regional level. The regions are listed [here](https://www.hhs.gov/about/agencies/iea/regional-offices/index.html). They are:

- 1: Connecticut, Maine, Massachusetts, New Hampshire, Rhode Island, and Vermont  
- 2: New Jersey, New York, Puerto Rico, and the Virgin Islands  
- 3: Delaware, District of Columbia, Maryland, Pennsylvania, Virginia, and West Virgina  
- 4: Alabama, Florida, Georgia, Kentucky, Mississippi, North Carolina, Suth Carolina, and Tennessee 
- 5: Illinois, Indiana, Michigan, Minnesota, Ohio, and Wisconsin  
- 6: Arkansas, Louisiana, New Mexico, Oklahoma, and Texas  
- 7: Iowa, Kansas, Missouri, and Nebraska  
- 8: Colorado, Montana, North Dakota, South Dakota, Utah, and Wyoming
- 9: Arizona, California, Hawaii, Nevada, American Samoa, Commonwealth of the Northern Mariana Islands, 
Federated States of Micronesia, Guam, Marshall Islands, and Republic of Palau  
- 10:Alaska, Idaho, Oregon, and Washington

## Accessing Data

### Downloading data

The dataset for [Vaccination Coverage among Adolescents](https://data.cdc.gov/Teen-Vaccinations/Vaccination-Coverage-among-Adolescents-13-17-Years/ee48-w5t6) was converted from a JSON object to and R object using `fromJSON` which is in the `jsonlite` package.  The dataset currently contains >25,000 lines (which is greater than the default for from JSON), so the function call needed an explicit limit.  

```{r, cache=TRUE}

vaccine_API <- fromJSON("https://data.cdc.gov/resource/ee48-w5t6.json?$limit=50000") %>% as_tibble

#Verify the number of rows is less than 50000 (that the limit was sufficient to retrieve the complete dataset).

nrow(vaccine_API)
vaccine_API

```

### Function 1 for HPV Vaccination Coverage: Regional level differences between genders with user modification of year 

The function `HPV_gender_eval_function` selects the rows related to HPV vaccination  at the regional level or national. The user may specify a specific year.  Possible values for years are each year of `2006` through `2022`.  The function selects rows corresponding to the coverage estimates for up to date vaccination of each of males and females in 13-17 year age range.

```{r, Function 1 for looking at HPV data-regional level }
HPV_gender_eval_function <- function(year_of_interest = "2022" ){
  
#Check input type
      if (!(year_of_interest %in% c('2006','2007','2008','2009','2010','2011','2012','2013','2014','2015',
                                    '2016','2017','2018','2019','2020','2021','2022'))) stop
        ("Year out of range, choose a year between 2006 and 2022 (inclusive).")
         
                            vaccine_API %>% 
                            filter(vaccine == "HPV" & geography_type == "HHS Regions/National") %>%  #Select vaccine=HPV and National/Regional Level
                            filter(year_season == year_of_interest) %>%                              #Filter Year based on user request
                            filter((dose == "Up-to-Date, Females")| (dose == "Up-to-Date, Males")) %>%  
                            filter(dimension == "13-17 Years")
}

```

A demonstration of the function (`HPV_gender_eval_function`) is given below.
```{r, demonstration of Function 1, message=TRUE, warning=TRUE}
demo_function_1 <- HPV_gender_eval_function(year_of_interest ="2021")
demo_function_1
```

A demonstration of the results from this function call transformed into a user-friendly table is given below.

```{r, table using data from Function 1, rows.print=11}
demo_function_1a <- demo_function_1 %>%select(geography, dose, coverage_estimate) %>% 
                   mutate (coverage_estimate=as.numeric(coverage_estimate)) %>% #to allow calculation of difference
                   pivot_wider(names_from=dose, values_from=(coverage_estimate)) 

#Rename columns for ease of reference in mutate 
colnames(demo_function_1a) <-c ("geography",  "cov_est_female", "cov_est_male")

#Add a column for difference between genders
demo_function_1a <- mutate(demo_function_1a, "Diff"=cov_est_female-cov_est_male)
demo_function_1a


```

*Considered on a national level, the difference between male and female vaccination rates is less than 5%. But in Regions 2 and 5, the  difference seems "higher" at 9%, suggesting this  might be worth further investigation.
``{r}

### Function 2 for HPV Vaccination Coverage: State level differences between genders with user modification of year 

The function `HPV_gender_eval_function_state` selects the rows related to HPV vaccination  at the state/territory level . It makes use of the state.name dataset built into R combined with the territories with entries in the dataset (District of Columbia, Guam, Puerto Rico, and U.S. Virgin Islands).  Additionally The user may specify a specific year.  Possible values for years are each year of `2006` through `2022`.  The function selects rows corresponding to the coverage estimates for up to date vaccination of each of males and females in 13-17 year age range.

```{r, Function 2 for looking at HPV data-state level }
HPV_gender_eval_function_state <- function(year_of_interest = "2022" ){
  
#Check input type
#      if (!(year_of_interest %in% c('2006','2007','2008','2009','2010','2011','2012','2013','2014','2015',
#                                    '2016','2017','2018','2019','2020','2021','2022'))) stop
#       ("Year out of range, choose a year between 2006 and 2022 (inclusive).")
         
                            vaccine_API %>% 
                            filter((vaccine == "HPV") & 
                                     (geography %in% state.name | 
                                        geography %in% c("District of Columbia", "Guam", "Puerto Rico", "U.S. Virgin Islands","United States"))) %>%
                            filter(year_season == year_of_interest) %>%                              #Filter Year based on user request
                            filter((dose == "Up-to-Date, Females")| (dose == "Up-to-Date, Males")) %>%  
                            filter(dimension == "13-17 Years")
}

```


A demonstration of the function (`HPV_gender_eval_function_state`) is given below.
```{r, demonstration of Function 2,error=TRUE,  message=TRUE, warning=TRUE}
demo_function_2 <- HPV_gender_eval_function_state(year_of_interest ="2021")
demo_function_2
```

A demonstration of the results from this function call transformed into a user-friendly table is given below.

```{r, table using data from Function 2}
demo_function_2a <- demo_function_2 %>%select(geography, dose, coverage_estimate)%>% 
                   mutate (coverage_estimate=as.numeric(coverage_estimate)) %>%
                   pivot_wider(names_from=dose, values_from=coverage_estimate)

#Rename columns for ease of reference in mutate 
colnames(demo_function_2a) <-c ("geography",  "cov_est_female", "cov_est_male")

#Add a column for difference between genders
demo_function_2a <- mutate(demo_function_2a, "Diff"=cov_est_female-cov_est_male)
demo_function_2a
```

*States that have more than a 15% difference (absolute) in coverage rates (with a higher percent of females being vaccinated) include New Jersey, Tennessee, Virginia, and West Virginia.  At the other extreme, Kentucky has more than a 15% difference (absolute) in coverage with a higher percent of males vaccinated.  The quick look at the data suggests that further study of the gender differences in HPV vaccine coverage correlated with region might  be interesting.*  



### Function 3 for  Vaccination Coverage for Vaccines other than HPV: Regional Level Differences 

The function `vaccines_regional_eval_function` selects the rows related to  vaccination with vaccines other than HPV at the regional level.  Varicella data is also being excluded because the compliance data is separated into entries for teens with just vaccinations, those with a history of the disease and those with both. 

The user can select a `year_of_interest` to use in the analysis.  Possible values for years are each year of `2006` through `2022`.  The user may also select.  A default value of 2022 is provided for `year_of_interest`

The user can select a `vaccine_of_interest`.  Possible values include:

- "Meningococcal Conjugate"  
- "Td or Tdap"  
- "Tdap"  
- "Hep A"  
- "MMR"  
- "HepB" 

A default value of NULL is provided for `vaccine_of_interest` this will return data for all rows with the vaccines equal to any in the above list.

-

```{r, Function 3 for looking at data for vaccines other than HPV data-regional level  }
vaccines_regional_eval_function <- function(vaccine_of_interest = NULL, year_of_interest = "2022" ){
  
    #Check input type
    if (!(year_of_interest %in% c('2006','2007','2008','2009','2010','2011','2012','2013','2014','2015',
                                    '2016','2017','2018','2019','2020','2021','2022'))) stop
        ("Year out of range, choose a year between 2006 and 2022 (inclusive).")
         
    filtered_API <- vaccine_API %>% 
    filter(vaccine != "HPV" & vaccine != "Varicella"          #Select vaccine=not HPV or Varicella
           & geography_type == "HHS Regions/National") %>%    #Select National/Regional Level
    filter(year_season == year_of_interest) %>%               #Filter Year based on user request
    filter (dimension =="13-17 Years")%>%                            #Filter age to allow comparison between vaccines
    mutate(coverage_estimate = as.numeric(coverage_estimate)) #Recast coverage_estimate to numeric 

    # filter for vaccine_of_interest if it is present, otherwise return all
    if (!is.null(vaccine_of_interest)) {
        filtered_API <- filtered_API %>% 
          filter(vaccine == vaccine_of_interest)
    }
    return(filtered_API)
}

```

A demonstration of the function (`vaccines_regional_eval_function`) is given below.

```{r, demonstration of Function 3 , message=TRUE, warning=TRUE}
demo_function_3 <- vaccines_regional_eval_function(year_of_interest = "2021")
```

Presentation of the results from this function call transformed into a user-friendly table is given below.

```{r, table using data from Function 3, rows.print=11}
demo_function_3a <- demo_function_3 %>%
                    select(geography, vaccine, coverage_estimate)%>% 
                    pivot_wider(names_from=vaccine, values_from=coverage_estimate)
demo_function_3a


```

*Based on this function call, it appears that vaccination coverage for Hepatitis A is lower than the other vaccines.  
[Immunization requiredments for secondary school](https://www.immunize.org/laws/hepa.asp) only include Hepatitis A in about half the states. This might be related.  Looking at the state data should include could show if lower rates of vaccination for Hepatitis A are observed only in states without a vaccination requirement. 


### Function 4 for  Vaccination Coverage for Vaccines other than HPV: State Level Differences 

The function `vaccines_state_eval_function` selects the rows related to  vaccination with vaccines other than HPV at the state level.  Varicella data is also being excluded because the compliance data is separated into entries for teens with just vaccinations, those with a history of the disease and those with both. 

The user can select a `year_of_interest` to use in the analysis.  Possible values for years are each year of `2006` through `2022`.  A default value of 2022 is provided for `year_of_interest`

The user can select a `vaccine_of_interest`.  Possible values include:

- "Meningococcal Conjugate"  
- "Td or Tdap"  
- "Tdap"  
- "Hep A"  
- "MMR"  
- "HepB" 

A default value of NULL is provided for `vaccine_of_interest` this will return data for all rows with the vaccines equal to any in the above list.

The user can select a `state_or_territory_of_interest`. Possible values include the fifty states as well as the territories with data present in the dataset (District of Columbia, Guam, Puerto Rico, and U.S. Virgin Islands.) A default value of NULL is provided for `state_of_interest` this will return data for all the states and territories.

```{r, Function 4 for looking at data for vaccines other than HPV or varicella data-state level  }
vaccines_state_eval_function <- function(vaccine_of_interest = NULL, state_of_interest=NULL, year_of_interest = "2022" ){
  
    #Check input type
    if (!(year_of_interest %in% c('2006','2007','2008','2009','2010','2011','2012','2013','2014','2015',
                                    '2016','2017','2018','2019','2020','2021','2022'))) 
       {stop
        ("Year out of range, choose a year between 2006 and 2022 (inclusive).")}
         
    filtered_API <- vaccine_API %>% 
    filter(vaccine != "HPV" & vaccine != "Varicella" &         #Select vaccine=not HPV or Varicella
          (geography %in% state.name | 
           geography %in% c("District of Columbia", "Guam", "Puerto Rico", "U.S. Virgin Islands"))) %>%
    filter(year_season == year_of_interest) %>%               #Filter Year based on user request
    filter (dimension =="13-17 Years")%>%                     #Filter age to allow easy comparison between vaccines
    mutate(coverage_estimate = as.numeric(coverage_estimate)) #Recast coverage_estimate to numeric 

    # filter for vaccine_of_interest if it is present, otherwise return all
    if (!is.null(vaccine_of_interest)) {
        filtered_API <- filtered_API %>% 
          filter(vaccine == vaccine_of_interest)
    }

    # filter for state_of_interest if it is present, otherwise return all
    if (!is.null(state_of_interest)) {
        filtered_API <- filtered_API %>% 
          filter(geography == state_of_interest)
    }

    return(filtered_API)
}
```

A demonstration of the function (`vaccines_state_eval_function`) is given below.

```{r, demonstration of Function 4 , message=TRUE, warning=TRUE}
demo_function_4 <- vaccines_state_eval_function(year_of_interest = "2021", state_of_interest="North Carolina")
demo_function_4
```
